Supplementary Information

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
minor updates

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)
n_tested=37 #same for all metrics
n_best_model_RW=30
n_best_model_BAI=31
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.3%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

@@

Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by this approach.

@@

Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
*(Discuss how this fits with regional to global-scale studies)*

While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), the approach differs in some substantive ways.
First, in determining the most important climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

@@

**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`).** 
Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
However, within these forest settings, the majority of species exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including three in this analysis (BCNM, HKK, SCBI), where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al. 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].

@@

Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

@@

It is notable that growth rates are changing directionally for most species/ sites.

>>>>

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Jan6)
n_tested=37 #same for all metrics
n_best_model_RW=30
n_best_model_BAI=31
n_best_model_AGB=30
n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.3%
n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
percent_sig_neg_trend_of_analyzed = n_best_model_neg_RW=n_best_model_RW/ n_tested*100
#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,

@@

Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
*(Discuss how this fits with regional to global-scale studies)*

While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the most important climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

@@

**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`).** 
Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern. *).
However, within these forest settings, the majority of species exhibited initially low, but increasing, $RW$.
This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including three in this analysis (BCNM, HKK, SCBI), where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015]. 
While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al. 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].

@@

Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed.
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

@@

It is notable that growth rates are changing directionally for the majority (`r percent_sig_neg_trend_of_analyzed`%) of site-species combinations that could be analyzed (Fig. `r year_fig`).

### 2021-01-12 doc/manuscript/MEE_manuscript_SI.Rmd
minor updates

SC_1970break_legend = "Figure 29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."

@@

![](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=70% }

@@

![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=70% }

\newpage
## `r RW_interactions_all_legend`

>>>>

SC_1970break_legend = "Figure S29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."

@@

![ `r `](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=57% }

@@

![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }

## `r RW_interactions_all_legend`
